Imprecise Datapath Design: An Overclocking Approach

### 2013-05-29 TRETS2013/TRETS_overclocking.tex

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. We perform our initial analysis based on the assumption that the carry propagation delay of each FA is a constant value $\mu$, which is a combination of logic delay and routing delay within an FPGA. Hence the critical path delay of the RCA is given by~(\ref{Eq:RCA_delay}), as shown in Fig.~\ref{FPGA adder}.
%
\begin{eqnarray}\label{Eq:RCA_delay}
	\mu_{RCA}=n\cdot\mu
\end{eqnarray}

% $\mu_{RCA}=n\mu$, as shown in~. 
For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

@@

In the following sections, we consider two methods that would allow the circuit to run at a frequency higher than $1/{T_S}$. The first is a traditional circuit design approach where operations occur without timing violations. To this end, the operand word-length is truncated in order to meet the timing requirement. This process results in truncation or roundoff error. In our proposed new scenario, circuits are implemented with greater word-length, but are clocked beyond the safe region so that timing violations sometimes occur. This process generates ``overclocking error''.

\subsection{Probabilistic Model of Truncation Error}\label{section_RCA_TruncationError}
For ease of discussion, we assume that the input to our circuit is a fixed point number scaled to lie in the range $[-1,1)$. For our initial analysis, we assume every bit of each input is uniformly and independently generated. However, this assumption will be relaxed in Section~\ref{Section_Experiments} where the predictions are verified using real image data. The errors at the output are evaluated in terms of the absolute value and the probability of their occurring. These two metrics are combined as the error expectation.

If the input signal of a circuit is $k$ bits, truncation error occurs when the input signal is truncated from $k$ bits to $n$ bits. Under this premise, the mean value of the truncated bits at signal input ($E_{Tin}$) is given by (\ref{ET at single input}).
%
\begin{eqnarray}\label{ET at single input}
%\small
%\scriptsize
%  \begin{split}
    E_{Tin} = \frac{1}{2}\sum_{i=n+1}^{k}2^{-i} = 2^{-n-1}-2^{-k-1}
%  \vspace{-1ex}
%  \end{split}
\end{eqnarray}

Since we assume there are two mutually independent inputs to the RCA, the overall expectation of truncation error for the RCA is given by (\ref{TruncationError}).
%
\begin{eqnarray}\label{TruncationError}
%\small
%\footnotesize
  E_T=\left\{
    \begin{matrix}
        2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
        0, & \textrm{otherwise}
    \end{matrix}
  \right.
\end{eqnarray}
%\normalsize

\subsection{Probabilistic Model of Overclocking Error}
\subsubsection{Generation of Overclocking Error}\label{subsub:Generation of Overclocking Error}
For a given $T_S$, the maximum length of error-free carry propagation is described by~(\ref{Max.length of carry chain}), where $f_S$ denotes the sampling frequency.
%
\begin{eqnarray}\label{Max.length of carry chain}
    b := \left\lceil \frac{T_S}{\mu} \right\rceil= \left\lceil \frac{1}{\mu\cdot f_S}\right\rceil
\end{eqnarray}

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$, according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item If $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item If $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $S_i=0$;
   \item If $A_i=B_i$, the current carry chain annihilates at bit $i$, and $S_i=1$.
 \end{itemize}

 \subsubsection{Absolute Value of Overclocking Error}
For an $n$-bit RCA, let $C_{tm}$ denote the carry chain generated at bit $S_t$ with the length of $m$ bits. For a certain $f_S$, the maximum length of error-free carry propagation, $b$, is determined through (\ref{Max.length of carry chain}). The presence of overclocking error requires $m>b$. Since the length of carry chain cannot be greater than $n$, parameters $t$ and $m$ are bounded by (\ref{t_RCA}) and (\ref{m_RCA}):
%
\begin{align}
%\small
%\footnotesize
%\scriptsize
  \label{t_RCA}  0&\leq t \leq n-b\\
    b&<m\leq n+1-t      \label{m_RCA}
%    \vspace{-1ex}
\end{align}

For $C_{tm}$, correct results will be generated from bit $S_t$ to bit $S_{t+b-1}$. Hence the absolute value of error seen at the output, normalized to the MSB ($2^n$), is given by (\ref{Output Error}), where $\hat{S_i}$ and $S_i$ denote the actual and error-free output of bit $i$ respectively.
%
\begin{eqnarray}\label{Output Error}
%\small
%\footnotesize
    e_{tm}=\frac{\left|\sum_{i=t+b}^{n}(S_i-\hat{S_i})\cdot 2^i\right|}{2^n}
%    \vspace{-1ex}
\end{eqnarray}

$S_i$ and $\hat{S_i}$ can be determined using the equations from the previous statements in Section~\ref{subsub:Generation of Overclocking Error}. In the error-free case, the carry will propagate from bit $S_t$ to bit $S_{t+m-1}$, and we will obtain $S_{t+b}=S_{t+b+1}=\cdots=S_{t+m-2}=0$ for carry propagation, and $S_{t+m-1}=1$ for carry annihilation. However, when a timing violation occurs, the carry will not propagate through all these bits. Substituting these values into (\ref{Output Error}) yields (\ref{etm}). Interestingly, the value of overclocking error has no dependence on the length of carry chain $m$.
%
\begin{eqnarray}\label{etm}
%\small
%\footnotesize
%  \begin{split}
    e_{tm} = \frac{\left|2^{t+m-1}-2^{t+m-2}-\dots-2^{t+b}\right|}{2^n} = 2^{t+b-n}
%  \end{split}
\end{eqnarray}

\subsubsection{Probability of Overclocking Error}
The carry chain $C_{tm}$ occurs when there is a carry generated at bit $t$, a carry annihilated at bit $t+m-1$ and the carry propagates in between. Consequently, its probability $P_{tm}$ is given by (\ref{Ptm_RCA}).
%
\begin{eqnarray}\label{Ptm_RCA}
%\small
%\footnotesize
  P_{tm}=P_{(A_t=B_t=1)}P_{(A_{t+m-1}=B_{t+m-1})}\cdot \prod_{i=t+1}^{t+m-2}P_{(A_i\neq B_i)}
%  \vspace{-1ex}
\end{eqnarray}

Under the assumption that $A$ and $B$ are mutually independent and uniformly distributed, we have $P_{(A_i=B_i=1)}=1/4$, $P_{(A_i\neq B_i)}=1/2$ and $P_{(A_i=B_i)}=1/2$, so $P_{tm}$ can be obtained by~(\ref{Ptm2}). Note that~(\ref{Ptm2}) takes into account the carry annihilation always occurs when $t+m-1=n$.
%
\begin{eqnarray}\label{Ptm2}
%\small
%\footnotesize
    P_{tm}=\left\{\begin{array}{ll}
      (1/2)^{m+1} & \textrm{if $t+m-1<n$}\\
      (1/2)^{m} & \textrm{if $t+m-1=n$}
    \end{array} \right.
%    \vspace{-1ex}
\end{eqnarray}

\subsubsection{Expectation of Overclocking Error}
Expectation of overclocking error can be expressed by (\ref{Eo_exp}).
%
\begin{eqnarray}\label{Eo_exp}
%\small
%\footnotesize
%\scriptsize
    %E_O=\sum_{t=0}^{n-b}\sum_{m=b+1}^{n-t+1}P_{tm}\cdot e_{tm}
    E_O=\sum_{t}\sum_{m}P_{tm}\cdot e_{tm}
%    \vspace{-1ex}
\end{eqnarray}

Using $P_{tm}$ and $e_{tm}$ from (\ref{etm}) and (\ref{Ptm2}) respectively, $E_O$ can be obtained by (\ref{Eo}).
%
\begin{eqnarray}\label{Eo}
%\small
%\footnotesize
      E_O=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
      \right.
%      \vspace{-1ex}
\end{eqnarray}

\subsection{Comparison between Two Scenarios}\label{section_RCA_Comparison}
In the traditional scenario, the word-length of RCA must be truncated, using $n=b-1$ bits, in order to meet a given $f_S$. The error expectation is then given by (\ref{Eo_trad_RCA}).
%
\begin{eqnarray}\label{Eo_trad_RCA}
%\small
%\vspace{-1ex}
%\footnotesize
%\scriptsize
  E_{trad}=2^{-b+1}-2^{-k}
  \vspace{-1ex}
\end{eqnarray}

Overclocking errors are allowed to happen in the second scenario, therefore the word-length of RCA is set to be equal to the input word-length, that is, $n=k$. Hence we obtain (\ref{Eo_new_RCA}) according to (\ref{Eo}).
%
\begin{eqnarray}\label{Eo_new_RCA}
%\vspace{-1ex}
%\small
%\footnotesize
%\scriptsize
  E_{new}=2^{-b}-2^{-k-1}
%  \vspace{-1ex}
\end{eqnarray}

Comparing (\ref{Eo_new_RCA}) and (\ref{Eo_trad_RCA}), we have (\ref{Comparison RCA}). This equation indicates that by allowing timing violations, the overall error expectation of RCA outputs drops by a factor of 2 in comparison to traditional scenario. This provides the first hint that our approach is useful in practice.
%
\begin{eqnarray}\label{Comparison RCA}
%\vspace{-1ex}
%\small
%\footnotesize
%\scriptsize
  \frac{E_{new}}{E_{trad}}=\frac{2^{-b}-2^{-k-1}}{2^{-b+1}-2^{-k}}=\frac{1}{2}
\end{eqnarray}

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
Since the delay of RCA is determined by the length of carry chain, alternative adder architectures, such as carry select adder (CSA) have been proposed to boost performance. For CSA, the carry chain is divided into multiple overlapped sections to increase the operating speed, as shown in Fig.~\ref{Fig_CSA_Structure}. There are multiple stages within a CSA. Each stage contains two RCAs and two multiplexers, as seen in Fig.~\ref{Fig_CSA_SingleStage}. For a given input, two additions are performed simultaneously where the carry input is zero and one respectively. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \caption{The structure diagram of CSA.}
\end{figure}

\subsection{Timing Models for Carry Select Adder}
We initially model the CSA in order to understand the relationship between the operating frequency and maximum word-length of the CSA. This information can then be employed to determine the truncation error based on the models presented in Section~\ref{section_RCA_TruncationError}.

In our analysis, the stage delay of stage $i$ in the CSA refers to the combination of the $i$-bit carry propagation delay and the delay of multiplexing the carry output. For a CSA with $s$ stages~($s\geqslant 2$), let the stage delay be denoted by $d_{s-1},~\dots,~d_{0}$, where $d_{s-1}$ and $d_{0}$ represent the delay of the most significant and the least significant stages, respectively. We still follow the aforementioned assumption that the critical path delay of the CSA is due to carry propagation and multiplexing the carry output, instead of generating the sum outputs. It should be noted that unlike other stages, the least significant stage of the CSA is only built by one RCA without multiplexers, since it is directly driven by the carry input. Hence we can obtain the delay of the $i^{th}$ stage as presented in~(\ref{CSA_SingleStageDelay}), where $\mu_{c}$, $\mu_{mux}$ and $n_i$ denote the delay of $1$-bit carry propagation, the delay of multiplexing and the word-length of the $i^{th}$ stage of the CSA, respectively.
%
\begin{eqnarray}\label{CSA_SingleStageDelay}
  d_i=\left\{
    \begin{matrix}
      n_i\cdot \mu_{c}+(s-i)\cdot\mu_{mux}, &\textrm{if $i\in\left[1,s-1\right]$}\\
      n_0\cdot \mu_c + (s-1)\cdot\mu_{mux}, & \textrm{if $i=0$}
    \end{matrix}
  \right.
\end{eqnarray}

Under the timing-driven design environment, the delay of each stage of CSA is equalized in order to achieve the fastest operation, as presented in~(\ref{CSA_DelayEqual}).
%
\begin{eqnarray}\label{CSA_DelayEqual}
  d_{s-1}=d_{s-2}=\cdots=d_{0}
\end{eqnarray}

In this case, combining~(\ref{CSA_SingleStageDelay}) and (\ref{CSA_DelayEqual}) yields $n_i$, which is represented by the word-length of the most significant stage $n_{s-1}$, in (\ref{CSA_DelayRep}).
%
\begin{eqnarray}\label{CSA_DelayRep}
 n_i=\left\{
	\begin{matrix}
	  n_{s-1}-(s-1-i)\cdot\frac{\mu_{mux}}{\mu_c}, & \textrm{if $i\in\left[1,s-1\right]$}\\
	  n_{s-1}-(s-2)\cdot\frac{\mu_{mux}}{\mu_c}, &\textrm{if $i=0$}
	\end{matrix}
    \right.
\end{eqnarray}

Sum up $n_i$ to give the total word-length of the CSA in (\ref{CSA_WL}).
%
\begin{eqnarray}\label{CSA_WL}
%  \begin{split}
    n_{CSA} =\sum_{i=0}^{s-1}n_{i}= s\cdot n_{s-1}-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+1)(s-2)}{2}
%  \end{split}
\end{eqnarray}

Conventionally under a given frequency constraint, the word-length of RCA is truncated by using $n_{RCA}=b-1$ bits, where $b$ is determined by~(\ref{Max.length of carry chain}). Similarly for CSA, the word-length of each stage should be selected in order to satisfy~(\ref{TimingConstraint_CSA}).
%
\begin{eqnarray}\label{TimingConstraint_CSA}
  \begin{matrix}
    \forall i\in[0,s-1] &, & d_i\leqslant\frac{1}{f_s}
  \end{matrix}
\end{eqnarray}

Hence for the same frequency requirement, we can form the relationship between the delay of the most significant stage of CSA and the delay of RCA, as given by~(\ref{WL_MSB_CSA}),
%
\begin{eqnarray}\label{WL_MSB_CSA}
    \mu_c\cdot (b-1) = n_{s-1}\cdot\mu_c+\mu_{mux}
\end{eqnarray}

Substitute~(\ref{WL_MSB_CSA}) into (\ref{CSA_WL}) to replace $n_{s-1}$, we derive the representation of the word-length of CSA in terms of $b$, as presented in~(\ref{CSA_Timing}).
%
\begin{eqnarray}\label{CSA_Timing}
  n_{CSA}=s\cdot (b-1)-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+2)(s-1)}{2}
\end{eqnarray}

\subsection{Accuracy Benefits and Area Overhead in CSA}
We first verify the models for the RCA and CSA in terms of the maximum word-length under the given operating frequencies. For the CSA, the ratio $\mu_{mux}/\mu_c$ can be computed experimentally. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{CSA_SingleStageDelay}) is recorded with respect to different word-lengths. The total word-length of CSA can be predicted through~(\ref{CSA_Timing}). In addition, the maximum word-lengths of the 2-stage CSA and RCA are obtained experimentally by increasing $n_{CSA}$ and $n_{RCA}$ respectively until errors are observed at the output. The comparison between the modeled value and the empirical results is illustrated in Fig.~\ref{Fig_CSA2stage_Timing}. It can be seen that our models for both RCA and CSA match well with the experimental results.

>>>>

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. We perform our initial analysis based on the assumption that the carry propagation delay of each FA is a constant value $\mu_c$, which is a combination of logic delay and routing delay within an FPGA. Hence the critical path delay of the RCA is given by~(\ref{Eq:RCA_delay}), as shown in Fig.~\ref{FPGA adder}.
%
\begin{eqnarray}\label{Eq:RCA_delay}
	\mu_{RCA}=n\cdot\mu_c
\end{eqnarray}

% $\mu_{RCA}=n\mu$, as shown in~. 
For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, faster-than-rated sampling frequencies are applied such that $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

@@

In the following sections, we consider two methods that would allow the circuit to run at a frequency higher than $1/{T_S}$. The first is a traditional circuit design approach where operations occur without timing violations. To this end, the operand word-length is truncated in order to meet the timing requirement. This process would result in truncation or roundoff error. In our proposed new design scenario, circuits are implemented with greater word-length, but are clocked beyond the safe region so that timing violations sometimes occur. This process would generate ``overclocking error''.

\subsection{Probabilistic Model of Truncation Error}\label{section_RCA_TruncationError}
For the ease of discussion, we assume that the input to our circuit is a fixed point number scaled to lie in the range $[-1,1)$. In our initial analysis, we assume every bit of each input is uniformly and independently generated. However, this assumption will be relaxed in Section~\ref{Section_Experiments} where the predictions are verified using real image data. The errors at the output are evaluated in terms of the absolute value and the probability of their occurring. These two metrics are combined as the error expectation.

If the input signal of a circuit contains $k$ bits, truncation error occurs when the input signal is truncated from $k$ bits to $n$ bits. Under this premise, the mean value of the truncated bits at signal input ($E_{Tin}$) is given by (\ref{ET at single input}).
%
\begin{eqnarray}\label{ET at single input}
%\small
%\scriptsize
%  \begin{split}
    E_{Tin} = \frac{1}{2}\sum_{i=n+1}^{k}2^{-i} = 2^{-n-1}-2^{-k-1}
%  \vspace{-1ex}
%  \end{split}
\end{eqnarray}

Since we assume there are two mutually independent inputs to the RCA, the overall expectation of truncation error for the RCA is given by (\ref{Eq:TruncationError}).
%
\begin{eqnarray}\label{Eq:TruncationError}
%\small
%\footnotesize
  E_T=\left\{
    \begin{matrix}
        2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
        0, & \textrm{otherwise}
    \end{matrix}
  \right.
\end{eqnarray}
%\normalsize

\subsection{Probabilistic Model of Overclocking Error}
\subsubsection{Generation of Overclocking Error}\label{subsub:Generation of Overclocking Error}
For a given $T_S$, the maximum length of error-free carry propagation is described by~(\ref{Eq:Max_CClength}), where $f_S$ denotes the sampling frequency.
%
\begin{eqnarray}\label{Eq:Max_CClength}
    b := \left\lceil \frac{T_S}{\mu_c} \right\rceil= \left\lceil \frac{1}{\mu_c\cdot f_S}\right\rceil
\end{eqnarray}

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$ within an $n$-bit RCA ($0\leq i<n$), according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item Carry generation: $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item Carry propagation: $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $S_i=0$;
   \item Carry annihilation: $A_i=B_i$ (either $0$ or $1$), the current carry chain annihilates at bit $i$, and $S_i=1$.
 \end{itemize}

 \subsubsection{Absolute Value of Overclocking Error}
For an $n$-bit RCA, let $C_{tm}$ denote the carry chain generated at bit $S_t$ with the length of $m$ bits. For a certain $f_S$, the maximum length of error-free carry propagation, $b$, is determined through (\ref{Eq:Max_CClength}). The presence of overclocking error requires $m>b$. Since the length of carry chain cannot be greater than $n$, parameters $t$ and $m$ are bounded by (\ref{Eq:t_RCA}) and (\ref{Eq:m_RCA}):
%
\begin{align}
%\small
%\footnotesize
%\scriptsize
  \label{Eq:t_RCA}  0&\leq t \leq n-b\\
    b&<m\leq n+1-t      \label{Eq:m_RCA}
%    \vspace{-1ex}
\end{align}

For $C_{tm}$, correct results will be generated from bit $S_t$ to bit $S_{t+b-1}$. Hence the absolute value of error seen at the output, normalized to the MSB ($2^n$), is given by (\ref{Eq:Output_Error}), where $\hat{S_i}$ and $S_i$ denote the actual value and error-free value of outputs at bit $i$, respectively.
%
\begin{eqnarray}\label{Eq:Output_Error}
%\small
%\footnotesize
    e_{tm}=\frac{\left|\sum_{i=t+b}^{n}(S_i-\hat{S_i})\cdot 2^i\right|}{2^n}
%    \vspace{-1ex}
\end{eqnarray}

$S_i$ and $\hat{S_i}$ can be determined using the equations from the previous statements in Section~\ref{subsub:Generation of Overclocking Error}. In the error-free case, the carry will propagate from bit $S_t$ to bit $S_{t+m-1}$, and we will obtain $S_{t+b}=S_{t+b+1}=\cdots=S_{t+m-2}=0$ for carry propagation, and $S_{t+m-1}=1$ for carry annihilation. However, when a timing violation occurs, the carry will not propagate through all these bits. Hence we have $\hat{S}_{t+b}=\hat{S}_{t+b-1}=\cdots=\hat{S}_{t+m-2}=1$ for carry propagation, and $\hat{S}_{t+m-1}=0$ for carry annihilation. Substituting these values into (\ref{Eq:Output_Error}) yields (\ref{Eq:etm}). Interestingly as seen in~(\ref{Eq:etm}), the magnitude of overclocking error has no dependence on the length of carry chain $m$.
%
\begin{eqnarray}\label{Eq:etm}
%\small
%\footnotesize
%  \begin{split}
    e_{tm} = \frac{\left|2^{t+m-1}-2^{t+m-2}-\dots-2^{t+b}\right|}{2^n} = 2^{t+b-n}
%  \end{split}
\end{eqnarray}

\subsubsection{Probability of Overclocking Error}
The carry chain $C_{tm}$ occurs when there is a carry generated at bit $t$, a carry annihilated at bit $t+m-1$ and the carry propagates in between. Consequently, its probability $P_{tm}$ is given by (\ref{Eq:Ptm_RCA}).
%
\begin{eqnarray}\label{Eq:Ptm_RCA}
%\small
%\footnotesize
  P_{tm}=P_{(A_t=B_t=1)}P_{(A_{t+m-1}=B_{t+m-1})}\cdot \prod_{i=t+1}^{t+m-2}P_{(A_i\neq B_i)}
%  \vspace{-1ex}
\end{eqnarray}

Under the assumption that $A$ and $B$ are mutually independent and uniformly distributed, we have $P_{(A_i=B_i=1)}=1/4$, $P_{(A_i\neq B_i)}=1/2$ and $P_{(A_i=B_i)}=1/2$, so $P_{tm}$ can be obtained by~(\ref{Eq:Ptm2}). Note that~(\ref{Eq:Ptm2}) takes into account the carry annihilation always occurs when $t+m-1=n$. 
%Furthermore, (\ref{Eq:Ptm_RCA}) can be employed for any probability distribution of inputs, as long as the value of own.
%
\begin{eqnarray}\label{Eq:Ptm2}
%\small
%\footnotesize
    P_{tm}=\left\{\begin{array}{ll}
      (1/2)^{m+1} & \textrm{if $t+m-1<n$}\\
      (1/2)^{m} & \textrm{if $t+m-1=n$}
    \end{array} \right.
%    \vspace{-1ex}
\end{eqnarray}

Furthermore, it can be seen that the probabilistic models for the magnitude and probability of overclocking errors can be easily generalized for any input distributions, since (\ref{Eq:Output_Error}) and (\ref{Eq:etm}) are not related to the input distribution, while (\ref{Eq:Ptm_RCA}) can be employed as long as the value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ is known.

\subsubsection{Expectation of Overclocking Error}
Expectation of overclocking error can be expressed by (\ref{Eq:Eo_exp}).
%
\begin{eqnarray}\label{Eq:Eo_exp}
%\small
%\footnotesize
%\scriptsize
    %E_O=\sum_{t=0}^{n-b}\sum_{m=b+1}^{n-t+1}P_{tm}\cdot e_{tm}
    E_O=\sum_{t}\sum_{m}P_{tm}\cdot e_{tm}
%    \vspace{-1ex}
\end{eqnarray}

Using $P_{tm}$ and $e_{tm}$ from (\ref{Eq:etm}) and (\ref{Eq:Ptm2}) respectively, $E_O$ can be obtained in (\ref{Eq:Eo}).
%
\begin{eqnarray}\label{Eq:Eo}
%\small
%\footnotesize
      E_O=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
      \right.
%      \vspace{-1ex}
\end{eqnarray}

\subsection{Comparison between Two Scenarios}\label{section_RCA_Comparison}
In the traditional scenario, the word-length of RCA must be truncated, using $n=b-1$ bits, in order to meet a given $f_S$. The error expectation is then given by (\ref{Eq:Eo_trad_RCA}).
%
\begin{eqnarray}\label{Eq:Eo_trad_RCA}
%\small
%\vspace{-1ex}
%\footnotesize
%\scriptsize
  E_{trad}=2^{-b+1}-2^{-k}
  \vspace{-1ex}
\end{eqnarray}

Overclocking errors are allowed to happen in our proposed new scenario. Therefore the word-length of RCA is set to be equal to the input word-length, that is, $n=k$. Hence we obtain (\ref{Eq:Eo_new_RCA}) according to (\ref{Eq:Eo}).
%
\begin{eqnarray}\label{Eq:Eo_new_RCA}
%\vspace{-1ex}
%\small
%\footnotesize
%\scriptsize
  E_{new}=2^{-b}-2^{-k-1}
%  \vspace{-1ex}
\end{eqnarray}

Comparing (\ref{Eq:Eo_new_RCA}) and (\ref{Eq:Eo_trad_RCA}), we have (\ref{Eq:Comparison RCA}). This equation indicates that by allowing timing violations, the overall error expectation of RCA outputs drops by a factor of 2 in comparison to traditional scenario. This provides the first hint that our approach is useful in practice.
%
\begin{eqnarray}\label{Eq:Comparison RCA}
%\vspace{-1ex}
%\small
%\footnotesize
%\scriptsize
  \frac{E_{new}}{E_{trad}}=\frac{2^{-b}-2^{-k-1}}{2^{-b+1}-2^{-k}}=\frac{1}{2}
\end{eqnarray}

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
Since the delay of RCA is determined by the length of carry chain, alternative adder architectures, such as carry select adder (CSA), have been proposed to boost performance. In a CSA, the carry chain is divided into multiple overlapped stages, while each stage contains two RCAs and two multiplexers, as shown in Fig.~\ref{Fig_CSA_Structure} and Fig.~\ref{Fig_CSA_SingleStage}, respectively. For a given input, two additions are performed simultaneously within a single stage where the carry input is zero and one seperately. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive to implement. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \caption{The structure diagram of CSA.}
\end{figure}

\subsection{Timing Models for Carry Select Adder}
We initially model the CSA in order to understand the relationship between the operating frequency and the maximum word-length of the CSA. This information can then be employed to determine the truncation error based on the models presented in Section~\ref{section_RCA_TruncationError}.

In our analysis, the stage delay of stage $i$ in the CSA refers to the combination of the $i$-bit carry propagation delay and the delay of multiplexing the carry output. For a CSA with $s$ stages~($s\geqslant 2$), let the stage delay be denoted by $d_{s-1},~\dots,~d_{0}$, where $d_{s-1}$ and $d_{0}$ represent the delay of the most significant and the least significant stages, respectively. We still follow the aforementioned assumption that the critical path delay of the CSA is due to carry propagation and multiplexing the carry output, instead of generating the sum outputs. It should be noted that unlike other stages, the least significant stage of the CSA can be built only by one RCA without multiplexers, since it is directly driven by the carry input. Hence we can obtain the delay of the $i^{th}$ stage as presented in~(\ref{Eq:CSA_SingleStageDelay}), where $\mu_{c}$, $\mu_{mux}$ and $n_i$ denote the delay of $1$-bit carry propagation, the delay of multiplexing and the word-length of the $i^{th}$ stage of the CSA, respectively.
%
\begin{eqnarray}\label{Eq:CSA_SingleStageDelay}
  d_i=\left\{
    \begin{matrix}
      n_i\cdot \mu_{c}+(s-i)\cdot\mu_{mux}, &\textrm{if $i\in\left[1,s-1\right]$}\\
      n_0\cdot \mu_c + (s-1)\cdot\mu_{mux}, & \textrm{if $i=0$}
    \end{matrix}
  \right.
\end{eqnarray}

Under the timing-driven design environment, the delay of each stage of CSA is equalized in order to achieve the fastest operation, as presented in~(\ref{Eq:CSA_DelayEqual}).
%
\begin{eqnarray}\label{Eq:CSA_DelayEqual}
  d_{s-1}=d_{s-2}=\cdots=d_{0}
\end{eqnarray}

In this case, combining~(\ref{Eq:CSA_SingleStageDelay}) and (\ref{Eq:CSA_DelayEqual}) yields $n_i$, which is represented by the word-length of the most significant stage $n_{s-1}$, in (\ref{Eq:CSA_DelayRep}).
%
\begin{eqnarray}\label{Eq:CSA_DelayRep}
 n_i=\left\{
	\begin{matrix}
	  n_{s-1}-(s-1-i)\cdot\frac{\mu_{mux}}{\mu_c}, & \textrm{if $i\in\left[1,s-1\right]$}\\
	  n_{s-1}-(s-2)\cdot\frac{\mu_{mux}}{\mu_c}, &\textrm{if $i=0$}
	\end{matrix}
    \right.
\end{eqnarray}

Sum up $n_i$ to give the total word-length of the CSA in (\ref{Eq:CSA_WL}).
%
\begin{eqnarray}\label{Eq:CSA_WL}
%  \begin{split}
    n_{CSA} =\sum_{i=0}^{s-1}n_{i}= s\cdot n_{s-1}-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+1)(s-2)}{2}
%  \end{split}
\end{eqnarray}

Conventionally under a given frequency constraint, the word-length of RCA is truncated by using $n_{RCA}=b-1$ bits, where $b$ is determined by~(\ref{Eq:Max_CClength}). Similarly for CSA, the word-length of each stage should be selected in order to satisfy~(\ref{Eq:TimingConstraint_CSA}).
%
\begin{eqnarray}\label{Eq:TimingConstraint_CSA}
  \begin{matrix}
    \forall i\in[0,s-1] &, & d_i\leqslant\frac{1}{f_s}
  \end{matrix}
\end{eqnarray}

Hence for the same frequency requirement, we can form the relationship between the delay of the most significant stage of CSA and the delay of RCA, as given by~(\ref{Eq:WL_MSB_CSA}),
%
\begin{eqnarray}\label{Eq:WL_MSB_CSA}
    \mu_c\cdot (b-1) = n_{s-1}\cdot\mu_c+\mu_{mux}
\end{eqnarray}

Substitute~(\ref{Eq:WL_MSB_CSA}) into (\ref{Eq:CSA_WL}) to replace $n_{s-1}$, we derive the representation of the word-length of CSA in terms of $b$, as presented in~(\ref{Eq:CSA_Timing}).
%
\begin{eqnarray}\label{Eq:CSA_Timing}
  n_{CSA}=s\cdot (b-1)-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+2)(s-1)}{2}
\end{eqnarray}

\subsection{Accuracy Benefits and Area Overhead in CSA}
We first verify the models for the RCA and CSA in terms of the maximum word-length under the given operating frequencies. For the CSA, the ratio $\mu_{mux}/\mu_c$ can be computed experimentally. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{Eq:CSA_SingleStageDelay}) is recorded with respect to different word-lengths. The total word-length of CSA can be predicted through~(\ref{Eq:CSA_Timing}). In addition, the maximum word-lengths of the 2-stage CSA and RCA are obtained experimentally by increasing $n_{CSA}$ and $n_{RCA}$ respectively until errors are observed at the output. The comparison between the modeled value and the empirical results is illustrated in Fig.~\ref{Fig_CSA2stage_Timing}. It can be seen that our models for both RCA and CSA match well with the experimental results.
